Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-05-14 analysis/paper/paper.Rmd
Some preliminary discussion of results

To visualise this temporal trend more clearly, a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. To account for the temporal uncertainty associated with the sites, a simulation-based approach was also adopted [e.g. @crema2012; @baxter2016; @orton2017]. A LOESS curve was fit to the data for each simulation run, where the age of each site was drawn from the date ranges associated with the sites. For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while dates for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range. This simulation was repeated 1000 times, the results of which is visualed in . Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE, before remaining relatively stable until the end of the Mesolithic. It is also interesting to note that the variation in degree of curation is higher after 8000 BCE, potentially reflecting variation in site types and associated mobility patterns.

@@

@@

```{r simage, cache = TRUE, fig.cap = "Temporal variation in the curation index. The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their associated probability distributions given in Figure 1B. A LOESS curve has been fit to the distribution for each of the 1000 simulation runs. Each simulation is plotted with some transparency. Sites mentioned in the text are coloured.", fig.height = 10, fig.width = 12}

@@

The results of the CA does appear to align reasonably well with previous research that has been based on more informal analyses of artefact types, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time. However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'. The developments evident here do also most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, highlighted perhaps most clearly by the influence of the later Mesolithic axe production sites. Unpicking and aligning the significance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural significance of the employed units of analysis unclear.

Assuming for now that the curation index captures degree of mobility, the present data indicates two major modes with a generally higher degree of mobility until some time before 8000 BCE. This then drops and stabilises from around 7000 BCE for the rest of the Mesolithic. This pattern is evident in both the flint data and when all lithics are treated in aggregate. The variation in degree of curation when all lithics are considered could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE. The five sites that have values on the curation index below c. -0.25 could in this perspective have predominantly functioned as base camps within a logistic settlement pattern. That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @eigeland2014b; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility. This highlights a possible issue pertaining to raw-material variability, as the coarse non-flint material used for the production of axes generally results in a relatively large amount of waste per produced tool, possibly skewing the curation index when compared to assemblages dominated by flint. However, even if this renders it more problematic to directly align the values on the curation index with a continuum between a residental and logistic settlement system, the variation on the index after around 8000 BCE is still evident. Another case worth commenting on is Langangen Vestgård 1, which, on the grounds of a large number of artefacts and the possibile presence of a dwelling structure was argued to reflect a more stable site location in the original report [@melvold2014]. However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of lower mobility, could represent a remnant from one or a few longer stays that constitute a smaller fraction of the use-life of the site as a whole.

>>>>

@@

To visualise this temporal trend more clearly, a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g. @crema2012; @baxter2016; @orton2017]. A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites. For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while age for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range. This simulation was repeated 1000 times, the results of which is visualised in \@ref(fig:simage). Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE. This stabilises around 7000 BCE without any further major fluctuations for the rest of the Mesolithic. It is also interesting to note that the variation in degree of curation is higher after 8000 BCE, potentially reflecting variation in site activities and associated mobility patterns.

@@

```{r simage, cache = TRUE, fig.cap = "Temporal variation in the curation index for A) all lithics in aggregate, and B) only the flint data. The temporal uncertainty is handled by means of a simulation approach, where the site ages are drawn from their repspective age determination probability distributions given in Figure 1B. A LOESS curve has been fit to the distribution for each of the 1000 simulation runs. Each simulation run is plotted with some transparency. Sites mentioned in the text are coloured.", fig.height = 10, fig.width = 12}

@@

The results of the CA does appear to align well with previous research that has been based on more informal treatment of the assemblages as a whole, indicating that the employed artefact categories are capturing some behaviorally meaningful patterning that changes over time. However, as several authors have pointed out in relation to the application of CA for seriation purposes, 'time is not the only dimension'. The developments evident here do also most certainly correspond to more pervasive cultural change than a purely typo-chronological development of artefact morphology, highlighted perhaps most clearly by the influence of the later Mesolithic axe production sites. Unpicking and aligning the significance of these chronological patterns with any specific behavioural dimensions using the CA results is, however, another task entirely. This follows most clearly from the fact that for the most part we do not know what individual lithic objects in the assemblages has been used for, leaving the behavioural significance of the employed units of analysis unclear. The results of the CA can, however, been used in conjunction with the part of the analysis that has attempted to get at more specific behavioural dimensions, to nuance or explain discrepencies in htis data.

Assuming for now that the curation index captures degree of mobility, the present data indicates a generally higher degree of mobility until some time before 8000 BCE. This then drops and stabilises from around 7000 BCE for the rest of the Mesolithic. This pattern is evident in both the flint data and when all lithics are treated in aggregate. The variation in degree of curation when all lithics are considered, could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE. The five sites that have values on the curation index below c. -0.25 could in this perspective have predominantly functioned as base camps within a logistic settlement pattern. That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @eigeland2014b; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility. This highlights a possible issue pertaining to raw-material variability, as the coarse non-flint material used for the production of axes generally results in a relatively large amount of waste per produced tool, possibly skewing the curation index when compared to assemblages dominated by flint. Also referring back to the CA, the difference is most marked for the sites in the later part of the Mesolithic where non-flint material become more dominating parts of the assemblages. As can be seen on the plot only using the flint-data, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages. The patterns evident in the flint data could thus indicate that only the period from around 8000 to 7500 BCE has individual sites that have had more stable roles within a settlement system, where some predominantly have functioned as base-camps. The lower degree of variation in the proceeding and preceeding phases could reflect either a more uniform purpose of visits to the sites, or that each assemblage consitutes the aggregation of visits of different purposes that in effect cancel each other out. Given the overall higher expediency in the earliest part of the Mesolithic this could be in line with a residental mobility pattern. For the later Mesolithic, the high expediency on some sites when considering all lithics in aggregate does speak to variability in site use, but the nature of the non-flint material does leave the signficance of this pattern more difficult to directly align with the residential logistic continuum. A case also worth commenting on is Langangen Vestgård 1, which, on the grounds of a overall large number of artefacts and the possible presence of a dwelling structure was argued to reflect a more permanent site location in the original report [@melvold2014]. However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of longer stays, could represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole.
